A generalized definition of multidimensional item response theory parameters

### 2024-07-18 output/paper_draft/paper_M2PLM_parameters.Rmd
Defines test space before, where it should

The main objective of this paper is thus to provide a generalization of
the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim to provide a method for
graphically representing these items in the fashion of @ackerman_creating_1994
but taking the non-orthogonality of the coordinate axes into account.
To pursue these objectives, we first define the generalization of
the `r MODEL_ACRONYM` model and describe its scope of application,
which will require introducing some linear algebra concepts and results.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce three different versions,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next, we discuss their properties,
propose a graphical representation method based on the new formulation,
and present two examples of application of our results.
We round up with a discussion about
the applicability and generalization of our results.

@@

Therefore, an inner product $`r TEST_SPACE_INNER_PROD`$ with Gram matrix
$`r INNER_PROD_MATRIX_INV`$ can be defined, such that $`r INNER_PROD_DISCR_EQ`$
keeps the norm of $`r DISCR_VECTOR`$ invariant.

@@

Before presenting the results about the parameters,
we must make a few remarks regarding the vector space $`r TEST_SPACE`$.
As shown before, this space is an entity by itself,
distinct from the latent space $`r LATENT_SPACE`$.
Despite this, the dot product in the exponent of Equation
\@ref(eq:M2PL-formulation) induces a bijective relationship between
their two sets of respective bases.
Explicitly, if $`r TRAIT_VECTOR`$ is represented in basis $`r LS_BASIS`$
with inner product Gram matrix $`r INNER_PROD_MATRIX`$,
$`r DISCR_VECTOR`$ is represented in basis $`r TEST_SPACE_BASIS`$
with inner product Gram matrix $`r INNER_PROD_MATRIX_INV`$,
for any pair of bases $`r LS_BASIS`$ and $`r TEST_SPACE_BASIS`$.
Only when the two bases are orthonormal,
the two spaces share a common inner product (i.e., the dot product),
and thus can be mistaken.
However, it is worth noting that, although their representation
can be superimposed in this case,
they are still two different spaces. <!-- # TODO: To discussion? -->
Following @zhang_theoretical_1999, we shall refer to space $`r TEST_SPACE`$ as
the *test space*, as it is the space where
the items (i.e., the test elements) are represented.

@@

With that purpose in mind, we have defined two versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based parameter has the drawback of not fulfilling
the first of @reckase_discriminating_1991's properties, namely,
the scale invariance property.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
More importantly, the covariance-based version
of the multidimensional parameters is invariant to
changes of latent-space basis.
Once again, this is important in applications where
potentially different bases are involved.
Given these considerations,
we propose to adopt the covariance-based version of
the multidimensional parameters as the most general definition,
and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of being scale-invariant,
which is convenient for fidelity when representing the items.
However, it is worth noting that it has been defined ad-hoc
to fulfill this specific property.
Therefore, it lacks other desirable properties of the covariance-based version,
like being invariant to a latent space scaling transformation,
or yielding a diagonal latent covariance matrix in the corresponding
orthogonal latent space basis.
Therefore, we recommend its use for graphical representation purposes only.
As for the agnostic version, we recommend it only in
the (unlikely) case of having no information about the latent space structure.

One may consider that the formal derivations presented here are
overly complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
@reckase_difficulty_1985's and @reckase_discriminating_1991's procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
Such a procedure will of course yield identical results,
given the invariance of the parameters to a change of basis.
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Computing the parameters in the orthonormalized (or just orthogonalized) space,
by contrast, would require computing the cosines in
the original basis afterwards.

On the other hand, establishing the relationship between the original and
the orthonormalized basis has provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
On the other hand, the latent space,
where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix.
This result is both surprising and counterintuitive;
The MIRT literature suggests in several instances that
"the angle between the axes [of latent ability space] represents
the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
We hope our results help shed a clearer light on
this often misunderstood topic.

It is worth noting that a similar relationship exists in
the common factor literature
between the *primary factors* and Thurstone's *reference axes*
[@harman_modern_1970, see p. 291].
Nevertheless, one must not make the mistake of assimilating the test space and
the *common-factor space*.
Their parameterization, albeit related [@mcdonald_test_1999], is different,
and we do not mean to imply that the derivations made here are directly
generalizable to the common factor model.
Those relationships deserve further exploration on their own.

Finally, we must also warn against generalizing these results
to other IRT models without further consideration.
First, we have only considered the case of dichotomous items,
whereas non-cognitive applications often (and sometimes cognitive ones as well)
require modeling several response categories.
Although proposals such as the graded-scale [@samejima_estimation_1968] and
the nominal response [@bock_estimating_1972] model are closely related to
the 2-parameter logistic model, it would be hasty to assume that these
parameters generalize in any way to
the multidimensional versions of those models without a formal proof.
The same can be said from the multidimensional parameters of other models
for dichotomous items different from the `r MODEL_ACRONYM` model.
We expect this work to help set the foundations for investigating those
generalizations.
Hopefully, this will contribute to a better understanding
of the underlying multidimensional measurement theory, its implications,
and its interpretation.

>>>>

The main objective of this paper is thus to provide a generalization of
the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim to provide a method for
graphically representing these items in the fashion of @ackerman_creating_1994
but taking the non-orthogonality of the coordinate axes into account.
In order to pursue these objectives, we first define the generalization of
the `r MODEL_ACRONYM` model and describe its scope of application,
which will require introducing some linear algebra concepts and results.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce three different versions,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next, we discuss their properties,
propose a graphical representation method based on the new formulation,
and present two examples of application of our results.
We round up with a discussion about
the applicability and generalization of our results.

@@

Therefore, an inner product $`r TEST_SPACE_INNER_PROD`$ with Gram matrix
$`r INNER_PROD_MATRIX_INV`$ can be defined, such that $`r INNER_PROD_DISCR_EQ`$
keeps the norm of $`r DISCR_VECTOR`$ invariant.
Following @zhang_theoretical_1999, we shall refer to space $`r TEST_SPACE`$ as
the *test space*, as it is the space where
the items (i.e., the test elements) are represented.

@@

Before presenting the results about the parameters,
we must make a few remarks regarding the vector space $`r TEST_SPACE`$.
As shown before, this space is an entity by itself,
distinct from the latent space $`r LATENT_SPACE`$.
Despite this, the dot product in the exponent of Equation
\@ref(eq:M2PL-formulation) induces a bijective relationship between
their two sets of respective bases.
Explicitly, if $`r TRAIT_VECTOR`$ is represented in basis $`r LS_BASIS`$
with inner product Gram matrix $`r INNER_PROD_MATRIX`$,
$`r DISCR_VECTOR`$ is represented in basis $`r TEST_SPACE_BASIS`$
with inner product Gram matrix $`r INNER_PROD_MATRIX_INV`$,
for any pair of bases $`r LS_BASIS`$ and $`r TEST_SPACE_BASIS`$.
Only when the two bases are orthonormal,
the two spaces share a common inner product (i.e., the dot product),
and thus can be mistaken.
However, it is worth noting that, although their representation
can be superimposed in this case,
they are still two different spaces. <!-- # TODO: To discussion? -->

@@

With that purpose in mind, we have defined two versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based parameter has the drawback of not fulfilling
the first of @reckase_discriminating_1991's properties, namely,
that $`r MDISC_UNIDIM_EQ`$ if an item only measures dimension $`r DIM_INDEX`$.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
Once again, this is important in applications where
potentially different latent spaces are involved.
Given these considerations, we propose to adopt the covariance-based version of
the multidimensional parameters and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of preserving the metric of the parameters,
which is convenient for fidelity when representing the items.
Therefore, we recommend using the correlation-based version of the parameters
for graphical representation purposes only.

One may consider that the formal derivations presented here are overly
complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
@reckase_difficulty_1985's and @reckase_discriminating_1991's procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Moreover, strictly following @reckase_difficulty_1985's method in the original,
oblique space has also provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
On the other hand, the latent space,
where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix.
It is worth noting that a similar relationship exists in
the common factor literature
between the *primary factors* and Thurstone's *reference axes*
[@harman_modern_1970, see p. 291].
Nevertheless, one must not make the mistake of assimilating the test space and
the *common-factor space*.
Their parameterization, albeit related [@mcdonald_test_1999], is different,
and we do not mean to imply that the derivations made here are directly
generalizable to the common factor model.
Those relationships deserve further exploration on their own.

Finally, we must also warn against generalizing these results
to other IRT models without further consideration.
First, we have only considered the case of dichotomous items,
whereas non-cognitive applications often (and sometimes cognitive ones as well)
require modeling several response categories.
Although models such as the graded-scale [@samejima_estimation_1968] and
the nominal response [@bock_estimating_1972] model are closely related to
the 2-parameter logistic model, it would be hasty to assume that these
parameters generalize in any way to
the multidimensional versions of those models without a formal proof.
The same can be said from the multidimensional parameters of other models
for dichotomous items different from the `r MODEL_ACRONYM` model.
We expect this work to help set the foundations for investigating those
generalizations.
Hopefully, this will contribute to a better understanding
of the underlying measurement theory, its implications and its interpretation.
